---
title: "DebarcodeR Tutorial"
author: "Benjamin Reisman"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DebarcodeR Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # pipeline steps are slow; run interactively
)
```

## Overview

DebarcodeR automates the demultiplexing of fluorescent cell barcoded (FCB)
flow cytometry data.  In FCB, samples are labeled with combinations of
amine-reactive dyes at varying concentrations before pooling.  After
acquisition, cells must be "debarcoded" back to their original sample
of origin.

The DebarcodeR pipeline has four steps:

1. **Deskew** — correct for cell-size-dependent dye uptake using
   morphology channels (FSC, SSC) and optionally an internal standard
2. **Cluster** — estimate per-cell probabilities of belonging to each
   barcoding level using a Gaussian mixture model or natural breaks
3. **Assign** — discretely assign cells to levels using likelihood and
   ambiguity cutoffs
4. **EM optimize** *(optional)* — refine assignments with a multivariate
   GMM that accounts for correlations between barcoding channels

The output is an `fcbFlowSet` (a subclass of `flowSet`) with one
`flowFrame` per debarcoded sample, compatible with the full
flowCore/Cytoverse ecosystem.

```{r libs, eval = TRUE, message = FALSE}
library(DebarcodeR)
library(flowCore)
library(dplyr)
```

---

## Dataset

The `jurkatFCB` dataset contains 191,233 Jurkat T cells barcoded with:

- **Pacific Blue-A**: 8 levels  
- **Pacific Orange-A**: 6 levels  
- **APC-H7-A** (Alexa Fluor 750): 1 level (internal standard)

for a total of 48 samples.  The cells were run in separate FCS files
and concatenated, with `row` and `col` columns recording each cell's
original sample of origin.

```{r load-data, eval = TRUE}
data("jurkatFCB")
jurkatFCB
```

We recommend filtering on the internal standard channel (`APC-H7-A`)
to remove unresponsive cells before debarcoding:

```{r filter-is}
filter_is  <- expressionFilter(`APC-H7-A` > 2.5, filterId = "ISfilter")
jurkatFCB  <- Subset(jurkatFCB, filter_is)
```

---

## External standard

DebarcodeR corrects for dye uptake heterogeneity using an **external
standard**: a control population stained with a single level of each
barcoding dye.  Because the tutorial data was acquired sample-by-sample,
we can extract a single well to serve as the external standard.

```{r ext-std}
filter_std    <- expressionFilter(`row` == 1 & `col` == 1, filterId = "std")
jurkatFCB_std <- Subset(jurkatFCB, filter_std)
```

---

## Step 1: Deskew

`deskew_fcbFlowFrame` fits a regression model (default: MARS via the
`earth` package) to predict dye uptake from scatter and the internal
standard channel, then subtracts the predicted component from each cell.

Channel names must be original FCS parameter names (e.g.
`"Pacific Blue-A"`, not `"pacific_blue_a"`).  Including the internal
standard (`APC-H7-A`) as a predictor substantially improves correction.

```{r deskew}
fcb <- fcbFlowFrame(jurkatFCB)

predictors <- c("FSC-A", "SSC-A", "FSC-H", "SSC-H", "FSC-W", "SSC-W", "APC-H7-A")

fcb <- deskew_fcbFlowFrame(fcb,
                           uptake     = jurkatFCB_std,
                           channel    = "Pacific Blue-A",
                           predictors = predictors)

fcb <- deskew_fcbFlowFrame(fcb,
                           uptake     = jurkatFCB_std,
                           channel    = "Pacific Orange-A",
                           predictors = predictors)
fcb
```

---

## Step 2: Cluster

`cluster_fcbFlowFrame` fits a Gaussian mixture model (via `mixsmsn`)
to the deskewed values for each channel to estimate the probability of
each cell originating from each barcoding level.

```{r cluster}
fcb <- cluster_fcbFlowFrame(fcb, channel = "Pacific Blue-A",   levels = 8)
fcb <- cluster_fcbFlowFrame(fcb, channel = "Pacific Orange-A", levels = 6)
fcb
```

---

## Step 3: Multivariate EM optimization *(optional)*

`em_optimize` refines the univariate clustering by fitting a
multivariate Gaussian mixture model across all barcoding channels.
This accounts for sample-to-sample correlated uptake variation
and typically improves well-level separation.

```{r em}
fcb <- em_optimize(fcb, niter = 5)
```

---

## Step 4: Assign

`assign_fcbFlowFrame` converts the probability matrix to discrete
assignments.  Two cutoffs control how conservatively cells are assigned:

- **`likelihoodcut`** (default 8): a cell must be at least
  1/*k* as probable as the most likely cell in its population.
- **`ambiguitycut`** (default 0.02): a cell must have a less than 2%
  chance of originating from any other population.

Unassigned cells receive level 0 rather than being discarded.

```{r assign}
fcb <- assign_fcbFlowFrame(fcb,
                           channel       = "wells",
                           likelihoodcut = 12,
                           ambiguitycut  = 0.05)
```

---

## Step 5: Split and apply plate map

`getAssignments` extracts the discrete assignments as a list of factors,
and `split` divides the flowFrame into one flowFrame per barcode
combination.

```{r split}
assignments  <- getAssignments(fcb)
debarcoded_fs <- split(jurkatFCB, assignments)
debarcoded_fs
```

`apply_platemap` renames each sample to a human-readable well label
using a lookup table mapping level combinations to well identifiers.
The column names in the platemap must use cleaned names (lowercase with
underscores, as produced by `janitor::clean_names()`).

```{r platemap}
myplatemap <- data.frame(
  pacific_blue_a   = as.character(rep(1:8, times = 6)),
  pacific_orange_a = as.character(rep(1:6, each  = 8)),
  well             = paste0(rep(LETTERS[1:8], times = 6),
                            formatC(rep(1:6, each = 8), width = 2, flag = "0"))
)

debarcoded_fs <- apply_platemap(debarcoded_fs,
                                myplatemap,
                                prefix = "Jurkat_FCB_001")
sampleNames(debarcoded_fs)
```

---

## Writing FCS files

The debarcoded `fcbFlowSet` is a standard `flowSet` and can be written
to disk with `flowCore::write.flowSet`:

```{r write}
dir.create("debarcoded_output", showWarnings = FALSE)
write.flowSet(debarcoded_fs, outdir = "debarcoded_output")
```

---

## Further reading

- Reisman et al. (2021) *DebarcodeR: Automated demultiplexing for
  fluorescent cell barcoding.* Cytometry A.
  [doi:10.1002/cyto.a.24363](https://doi.org/10.1002/cyto.a.24363)
- Krutzik et al. (2006) *Fluorescent cell barcoding in flow cytometry
  allows high-throughput drug screening and signaling profiling.*
  Nature Methods.
