---
title: "DebarcodeR Tutorial"
author: "Benjamin Reisman"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{DebarcodeR Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

DebarcodeR automates the demultiplexing of fluorescent cell barcoded (FCB)
flow cytometry data.  In FCB, samples are labeled with combinations of
amine-reactive dyes at varying concentrations before pooling.  After
acquisition, cells must be "debarcoded" back to their original sample
of origin.

The DebarcodeR pipeline has four steps:

1. **Deskew** — correct for cell-size-dependent dye uptake using
   morphology channels (FSC, SSC) and optionally an internal standard
2. **Cluster** — estimate per-cell probabilities of belonging to each
   barcoding level using a Gaussian mixture model or natural breaks
3. **Assign** — discretely assign cells to levels using likelihood and
   ambiguity cutoffs
4. **EM optimize** *(optional)* — refine assignments with a multivariate
   GMM that accounts for correlations between barcoding channels

The output is an `fcbFlowSet` (a subclass of `flowSet`) with one
`flowFrame` per debarcoded sample, compatible with the full
flowCore/Cytoverse ecosystem.

```{r libs, message = FALSE}
library(DebarcodeR)
library(flowCore)
library(ggplot2)
library(scales)
```

## Dataset

The `jurkatFCB` dataset contains 24,000 Jurkat T cells (a representative
subsample of the full experiment) barcoded with:

- **Pacific Blue-A**: 8 levels
- **Pacific Orange-A**: 6 levels
- **APC-H7-A** (Alexa Fluor 750): 1 level (internal standard)

for a total of 48 samples.  The cells were run in separate FCS files
and concatenated, with `row` and `col` columns recording each cell's
original sample of origin.

```{r load-data}
data("jurkatFCB")
jurkatFCB
```

We recommend filtering on the internal standard channel (`APC-H7-A`)
to remove unresponsive cells before debarcoding:

```{r filter-is}
filter_is  <- expressionFilter(`APC-H7-A` > 2.5, filterId = "ISfilter")
jurkatFCB  <- Subset(jurkatFCB, filter_is)
```

The two barcoding channels form an 8x6 grid — one cluster per sample:

```{r plot-raw, fig.width = 5, fig.height = 5, message = FALSE, warning = FALSE}
set.seed(6183)
plot_raw <- as.data.frame(exprs(jurkatFCB))
ggplot(plot_raw, aes(x = `Pacific Blue-A`, y = `Pacific Orange-A`)) +
  geom_bin2d(bins = 150) +
  scale_fill_viridis_c(option = "A", trans = "sqrt", name = "Count") +
  labs(x = "Pacific Blue-A", y = "Pacific Orange-A",
       title = "Raw barcoding channels: 8 \u00d7 6 sample grid") +
  coord_fixed() +
  theme_classic()
```

## External standard

DebarcodeR corrects for dye uptake heterogeneity using an **external
standard**: a control population stained with a single level of each
barcoding dye.  `jurkatFCB_std` is a pre-extracted single-well subset of
`jurkatFCB` (row 1, column 1) that serves as the external standard for
this tutorial.

```{r ext-std}
data("jurkatFCB_std")
jurkatFCB_std
```

## Step 1: Deskew

`deskew_fcbFlowFrame` fits a regression model (default: MARS via the
`earth` package) to predict dye uptake from scatter and the internal
standard channel, then subtracts the predicted component from each cell.

Channel names must be original FCS parameter names (e.g.
`"Pacific Blue-A"`, not `"pacific_blue_a"`).  Including the internal
standard (`APC-H7-A`) as a predictor substantially improves correction.

```{r deskew, message = FALSE}
fcb <- fcbFlowFrame(jurkatFCB)

predictors <- c("FSC-A", "SSC-A", "FSC-H", "SSC-H", "FSC-W", "SSC-W", "APC-H7-A")

fcb <- deskew_fcbFlowFrame(fcb,
                           uptake     = jurkatFCB_std,
                           channel    = "Pacific Blue-A",
                           predictors = predictors)

fcb <- deskew_fcbFlowFrame(fcb,
                           uptake     = jurkatFCB_std,
                           channel    = "Pacific Orange-A",
                           predictors = predictors)
fcb
```

Deskewing removes the correlation between cell size and fluorescence
intensity, sharpening the separation between adjacent levels:

```{r plot-deskew, fig.width = 7, fig.height = 3.5, message = FALSE, warning = FALSE}
raw_pb      <- exprs(jurkatFCB)[, "Pacific Blue-A"]
deskewed_pb <- get_barcode_data(fcb, "Pacific Blue-A", "deskewing", "values")

deskew_long <- rbind(
  data.frame(step = "Before deskewing", value = raw_pb),
  data.frame(step = "After deskewing",  value = deskewed_pb)
)
deskew_long$step <- factor(deskew_long$step,
                            levels = c("Before deskewing", "After deskewing"))
ggplot(deskew_long, aes(x = value)) +
  geom_histogram(bins = 150, fill = "steelblue4", color = NA) +
  facet_wrap(~step, nrow = 1) +
  labs(x = "Pacific Blue-A", y = "Cell count",
       title = "Deskewing sharpens level separation") +
  theme_classic()
```

## Step 2: Cluster

`cluster_fcbFlowFrame` fits a Gaussian mixture model (via `mixsmsn`)
to the deskewed values for each channel to estimate the probability of
each cell originating from each barcoding level.

```{r cluster, message = FALSE, warning = FALSE}
fcb <- cluster_fcbFlowFrame(fcb, channel = "Pacific Blue-A",
                            levels = 8, subsample = 10000)
fcb <- cluster_fcbFlowFrame(fcb, channel = "Pacific Orange-A",
                            levels = 6, subsample = 10000)
fcb
```

The Gaussian mixture model identifies each level as a distinct component.
Cells are colored by their most probable level assignment:

```{r plot-cluster, fig.width = 6.5, fig.height = 4, message = FALSE, warning = FALSE}
prob_mat <- get_barcode_data(fcb, "Pacific Blue-A", "clustering", "probabilities")
level <- factor(colnames(prob_mat)[max.col(prob_mat)],
                levels = colnames(prob_mat))

cl_df <- data.frame(deskewed = deskewed_pb, level = level)
ggplot(cl_df, aes(x = deskewed, fill = level)) +
  geom_histogram(bins = 150, position = "stack", color = NA) +
  scale_fill_brewer(palette = "RdYlBu", name = "Level",
                    guide = guide_legend(reverse = TRUE)) +
  labs(x = "Pacific Blue-A (deskewed)", y = "Cell count",
       title = "Clustering: 8 levels identified (Pacific Blue-A)") +
  theme_classic()
```

## Step 3: Multivariate EM optimization *(optional)*

`em_optimize` refines the univariate clustering by fitting a
multivariate Gaussian mixture model across all barcoding channels.
This accounts for sample-to-sample correlated uptake variation
and typically improves well-level separation.

```{r em, message = FALSE}
fcb <- assign_fcbFlowFrame(fcb, channel = "Pacific Blue-A")
fcb <- assign_fcbFlowFrame(fcb, channel = "Pacific Orange-A")
fcb <- em_optimize(fcb, niter = 3)
```

## Step 4: Assign

`assign_fcbFlowFrame` converts the probability matrix to discrete
assignments.  Two cutoffs control how conservatively cells are assigned:

- **`likelihoodcut`** (default 8): a cell must be at least
  1/*k* as probable as the most likely cell in its population.
- **`ambiguitycut`** (default 0.02): a cell must have a less than 2%
  chance of originating from any other population.

Unassigned cells receive level 0 rather than being discarded.

```{r assign}
fcb <- assign_fcbFlowFrame(fcb,
                           channel       = "wells",
                           likelihoodcut = 12,
                           ambiguitycut  = 0.05)
```

After assignment, the 8x6 grid of samples is cleanly resolved in the
deskewed channel space (cells not meeting the cutoffs are excluded):

```{r plot-assign, fig.width = 5.5, fig.height = 5.5, message = FALSE, warning = FALSE}
asgn <- getAssignments(fcb)
deskewed_po <- get_barcode_data(fcb, "Pacific Orange-A", "deskewing", "values")
assigned <- asgn[["Pacific Blue-A"]] != "0" & asgn[["Pacific Orange-A"]] != "0"
well_id <- paste(asgn[["Pacific Blue-A"]], asgn[["Pacific Orange-A"]], sep = ".")

asgn_df <- data.frame(
  pb   = deskewed_pb[assigned],
  po   = deskewed_po[assigned],
  well = well_id[assigned]
)
n_wells <- length(unique(asgn_df$well))

ggplot(asgn_df, aes(x = pb, y = po, color = well)) +
  geom_point(size = 0.15, alpha = 0.4) +
  scale_color_manual(values = hue_pal()(n_wells), guide = "none") +
  labs(x = "Pacific Blue-A (deskewed)", y = "Pacific Orange-A (deskewed)",
       title = "Final assignments: 48 samples resolved") +
  coord_fixed() +
  theme_classic()
```

## Step 5: Split and apply plate map

`getAssignments` extracts the discrete assignments as a list of factors,
and `split` divides the flowFrame into one flowFrame per barcode
combination.

```{r split}
debarcoded_fs <- split(jurkatFCB, asgn)
debarcoded_fs
```

`apply_platemap` renames each sample to a human-readable well label
using a lookup table mapping level combinations to well identifiers.
The column names in the platemap must use cleaned names (lowercase with
underscores, as produced by `janitor::clean_names()`).

```{r platemap}
myplatemap <- data.frame(
  pacific_blue_a   = as.character(rep(seq_len(8), times = 6)),
  pacific_orange_a = as.character(rep(seq_len(6), each  = 8)),
  well             = paste0(rep(LETTERS[seq_len(8)], times = 6),
                            formatC(rep(seq_len(6), each = 8), width = 2, flag = "0"))
)

debarcoded_fs <- apply_platemap(debarcoded_fs,
                                myplatemap,
                                prefix = "Jurkat_FCB_001")
sampleNames(debarcoded_fs)
```

## Writing FCS files

The debarcoded `fcbFlowSet` is a standard `flowSet` and can be written
to disk with `flowCore::write.flowSet`:

```{r write, eval = FALSE}
dir.create("debarcoded_output", showWarnings = FALSE)
write.flowSet(debarcoded_fs, outdir = "debarcoded_output")
```

## Further reading

- Reisman et al. (2021) *DebarcodeR: Automated demultiplexing for
  fluorescent cell barcoding.* Cytometry A.
  [doi:10.1002/cyto.a.24363](https://doi.org/10.1002/cyto.a.24363)
- Krutzik et al. (2006) *Fluorescent cell barcoding in flow cytometry
  allows high-throughput drug screening and signaling profiling.*
  Nature Methods.

```{r session-info}
sessionInfo()
```
